---
title: "Predicting the Manner in which an Exercise was done using the Weight Lifting Exercises
 Dataset"
author: Grzegorz Chlebus
date: 11/06/2020
output: html_document
---

## Introduction
The Weight Lifting Exercises Dataset contains sensor measurements of six participants performing
 the unilateral dumbbell biceps curl in different manners (one correct four incorrect ones).
In this assignment I'll investigate whether the manner in which the exercise was performed can be
 predicted from the measurements.

```{r, include = FALSE}
require(caret)
require(corrplot)
```

## EDA
```{r}
pmlTraining <- read.csv(file = "./data/pml-training.csv")
pmlTesting <- read.csv(file = "./data/pml-testing.csv")
dim(pmlTraining)
```
Columns with mostly nan values and near zero variance are removed as they have none predictive
 power. Additionally, columns with values (e.g., row number, timestamps, user name) that shouldn't be used as predictors are removed.
```{r}
# remove columns containing mostly nan values
nanColumns <- which(colSums(sapply(pmlTraining, is.na)) > (0.95 * nrow(pmlTraining)))
sprintf("Dropping %d columns, which contain mostly nan values.", length(nanColumns))
pmlTraining <- pmlTraining[, -nanColumns]
pmlTesting <- pmlTesting[, -nanColumns]

# remove column with near zero variance
nearZeroVarColumns <- nearZeroVar(pmlTraining)
sprintf("Dropping %d columns, which have near zero variance.", length(nearZeroVarColumns))
pmlTraining <- pmlTraining[, -nearZeroVarColumns]
pmlTesting <- pmlTesting[, -nearZeroVarColumns]

# remove columns that we don't want to use as predictors, i.e., row number, timestamps, user name
# , num_window
pmlTraining <- pmlTraining[, -c(1:6)]
pmlTesting <- pmlTesting[, -c(1:6)]
```

Most of the features have low correlation with the other ones.
```{r}
correlation <- cor(pmlTraining[, !(names(pmlTraining) %in% c("classe"))])
corrplot(correlation, type = "upper", tl.cex = 0.5)
```

## Model Selection
Data split.
```{r}
set.seed(1234)
inTraining <- createDataPartition(pmlTraining$classe, p = .8, list = FALSE)
training <- pmlTraining[inTraining,]
testing <- pmlTraining[-inTraining,]
```

Let's train a knn classifier, which is a simple classification model.
```{r}
set.seed(1234)
trControl <- trainControl(method = "cv", number = 5)
knnModel <- train(classe ~ ., data = training, method = "knn", trControl = trControl)
confMat <- confusionMatrix(data = predict(knnModel, testing), reference = testing$classe)
confMat$overall["Accuracy"]
```
Let's train a random forest classifier, which is a more advanced classification algorithm, which
 is known for its good performance.
```{r}
set.seed(1234)
trControl <- trainControl(method = "cv", number = 5)
rfModel <- train(classe ~ ., data = training, method = "rf", verbose = FALSE, trControl = trControl)
confMat <- confusionMatrix(data = predict(rfModel, testing), reference = testing$classe)
confMat$overall["Accuracy"]
```
The random forest performs much better on the test set. Therefore, the random forest model is
 applied to the 20 test cases provided with the assignment.
```{r}
testPredictions <- predict(rfModel, pmlTesting)
testPredictions
```
